---
title: "AncientMetagenomeDir - Analysis"
output:
  html_document:
    df_print: paged
---

# Introduction

AncientMetagenomeDir is a repository listing published ancient metagenome (and 
related) samples. It contains critical metadata for ancient metagenome studies, 
and acts as a 'sign post' towards important data to help facilitate more robust
and efficient comparative data.

This page provides summary statistics of the current status of the directory.

# Libraries and Versions

```{r}
library(tidyverse)
library(scales)
library(lubridate)
library(maps)
library(patchwork)
sessionInfo()
```

# Preparation

## Functions

We will prepare some custom functions


### Loading

```{r}
## Load and standardise date across tables
load_thedir_data <- function(path, name) {
  read_tsv(path, col_types = cols()) %>%
    mutate(List = name) %>%
    select(List, everything())
}

```

### Publication Timeline

```{r}

stats_pub_timeline <- function(...) {
  x <- list(...)

  ## Get only relevent columns
  lapply(x, FUN = function(y) {
    select(
      y,
      List, publication_doi,
      publication_year
    ) %>% distinct()
  }) %>%
    bind_rows() %>%
    mutate(List = factor(List, levels = names(dir_colours)))
}

plot_pub_timeline <- function(dat) {
  ## Get range so we plot x-axis nicely
  spanning_years <- dat %>% summarise(min = min(publication_year), max = max(publication_year))

  ggplot(dat, aes(publication_year, fill = List)) +
    scale_fill_manual(values = dir_colours, guide = guide_legend(ncol = 1)) +
    scale_y_continuous(labels = scales::number_format(accuracy = 1)) +
    ylab("Number of publications") +
    xlab("Publication year") +
    geom_bar(bins = spanning_years$max - spanning_years$min, binwidth = 1) +
    scale_x_continuous(breaks = seq(spanning_years$min, spanning_years$max, 2)) +
    theme_classic() +
    theme(legend.position = "bottom") +
    facet_wrap(~List, ncol = 1) +
    labs(fill = NULL)
}
```

### Cumulative Samples Timeline

```{r}
stats_cumulative_timeline <- function(...) {
  ## Takes a list of AncientMetagenomeDir TSVs
  x <- list(...)

  ## Get only relevent columns
  dat <- lapply(x, FUN = function(y) {
    select(
      y,
      List, sample_name, publication_year,
      publication_year
    ) %>% distinct()
  }) %>%
    bind_rows() %>%
    mutate(List = factor(List, levels = names(dir_colours)))

  spanning_years <- dat %>%
    ungroup() %>%
    summarise(min = min(publication_year), max = max(publication_year))

  ## Make fake base table to ensure all years for all lists
  ## Currently manually defined
  base_table <- list(
    seq(spanning_years$min, spanning_years$max, 1),
    seq(spanning_years$min, spanning_years$max, 1),
    seq(spanning_years$min, spanning_years$max, 1)
  )

  names(base_table) <- levels(dat$List)

  base_table <- base_table %>%
    enframe(name = "List", value = "publication_year") %>%
    unnest(publication_year)

  dat <- dat %>%
    group_by(List, publication_year) %>%
    summarise(count = n())

  dat %>%
    right_join(base_table, by = c("List", "publication_year")) %>%
    replace_na(list(count = 0)) %>%
    arrange(List, publication_year) %>%
    mutate(List = factor(List, levels = names(dir_colours))) %>%
    group_by(List) %>%
    mutate(cumulative_sum = cumsum(count))
}


plot_cumulative_timeline <- function(x) {
  
  spanning_years <- list(min_year = min(x$publication_year), max_year = max(x$publication_year))

  ## Get range so we plot x-axis nicely
  ggplot(x, aes(publication_year, cumulative_sum, fill = List)) +
    geom_bar(stat = "identity") +
    scale_y_continuous(labels = scales::number_format(accuracy = 1)) +
    scale_x_continuous(breaks = seq(spanning_years$min_year, spanning_years$max_year, 2)) +
    theme_classic() +
    xlab("Publication year") +
    ylab("Number of samples (cumulative sum)") +
    scale_fill_manual(values = dir_colours, guide = guide_legend(ncol = 1)) +
    theme(legend.position = "bottom") +
    facet_wrap(~List, ncol = 1) +
    labs(fill = NULL)
  
}


```

### Map

```{r}
stats_map <- function(...){

x <- list(raw_environmental, raw_hostmetagenome, raw_hostsinglegenome)

dat <- lapply(x, FUN = function(y) {
  select(
    y,
    List, sample_name, geo_loc_name, latitude, longitude,
    publication_year
  ) %>% distinct()
}) %>%
  bind_rows() %>%
  mutate(List = factor(List, levels = names(dir_colours)))

dat <- dat %>%
  group_by(List, geo_loc_name, latitude, longitude) %>%
  summarise(count = n()) %>%
  mutate(List = factor(List, levels = names(dir_colours)))
}


plot_map <- function(dat){
  
  world_map <- map_data("world")
  
  ggplot() +
    geom_polygon(data = world_map, aes(x = long, y = lat, group = group), fill = "white", colour = "grey", size = 0.3) +
    geom_point(data = dat, aes(x = longitude, y = latitude, fill = List, size = count), shape = 21, alpha = 0.5) +
    theme_linedraw() +
    facet_wrap(~List, ncol = 1) +
    theme_classic() +
    scale_fill_manual(values = dir_colours) +
    theme(legend.position = "bottom", legend.direction="vertical") +
    labs(fill = "Sample Type", size = "Sample Count") +
    xlab("Longitude") +
    ylab("Latitude") +
    guides(fill = FALSE)
  
}
```

### Age Timeline

```{r}
stats_age_timeline <- function(...){
  x <- list(...)
  
  dat <- lapply(x, FUN = function(y) {
    select(
      y,
      List, sample_name, geo_loc_name, latitude, longitude, sample_age, publication_year
    ) %>%
      distinct()
  }) %>%
    bind_rows() %>%
    mutate(List = factor(List, levels = names(dir_colours)))
  
  return(dat)
}

# from https://stackoverflow.com/a/11054781
reverselog_trans <- function(base = exp(1)) {
  trans <- function(x) -log(x, base)
  inv <- function(x) base^(-x)
  trans_new(paste0("reverselog-", format(base)), trans, inv,
            log_breaks(base = base),
            domain = c(1e-100, Inf)
  )
}

# Plot histograms, with custom binning!
plot_figure_age_timeline <- function(X, dataset, nclass = "Sturges", log = TRUE, group_order){
  
  X <- X %>% filter(List == dataset) %>% ungroup()
  
  # We steal the breaks from the base histogram
  if (log){
    hist_data <- hist(log10(X$sample_age), plot = FALSE, breaks = nclass)
    scale_log = scale_x_reverse(breaks = c(2,3,4,5), labels = comma(c(10^2, 10^3, 10^4, 10^5)), limits = c(5, 2))
  } else {
    hist_data <- hist(X$sample_age, plot = FALSE, breaks = nclass)
    scale_log = scale_x_reverse()
  }
  
  # We calculate where we want to have the columns in the plot
  width <- mean(diff(hist_data$breaks))
  nudge <- -(width/2)
  hist_df <- tibble(counts = hist_data$counts, breaks = (hist_data$breaks[1:length(hist_data$breaks)-1]), List = dataset) %>%
    mutate(width = width, nudge = nudge, 
           List = factor(List, levels = group_order))
  
  # Let's plot
  plot <- ggplot(hist_df, aes(breaks, counts, fill = List)) +
    geom_col(color = "black", width = width, position = position_nudge(x = nudge)) +
    #scale_y_log10() +
    scale_log +
    theme_classic(base_size = 8) +
    scale_fill_manual(values = dir_colours, guide = guide_legend(ncol = 1, reverse = T)) +
    ylab("Samples (n)") +
    xlab("Years before present")  +
    theme(legend.position = "none",
          strip.background = element_blank())
  return(list(plot = plot, data = hist_df))
}

```


## Data Loading

We will load and analyse each list separately, as they contain slightly different
metadata depending on context.

```{r}
raw_hostmetagenome <- load_thedir_data("https://raw.githubusercontent.com/SPAAM-community/AncientMetagenomeDir/v21.09/ancientmetagenome-hostassociated/ancientmetagenome-hostassociated.tsv", "Host Associated Metagenome") 
raw_hostsinglegenome <- load_thedir_data("https://raw.githubusercontent.com/SPAAM-community/AncientMetagenomeDir/v21.09/ancientsinglegenome-hostassociated/ancientsinglegenome-hostassociated.tsv", "Host Associated Single Genome") 
raw_environmental <- load_thedir_data("https://raw.githubusercontent.com/SPAAM-community/AncientMetagenomeDir/v21.09/ancientmetagenome-environmental/ancientmetagenome-environmental.tsv", "Environmental Metagenome") 
```

## Design Assets

```{r}
dir_colours <- c(`Host Associated Metagenome` = "#73cff3",
            `Host Associated Single Genome` = "#d74182",
            `Environmental Metagenome` = "#2da46a")

```


# Publication Timelines

```{r, fig.width=5, fig.height=7}
figure_publication_time <- stats_pub_timeline(raw_hostmetagenome, raw_hostsinglegenome, raw_environmental) %>% plot_pub_timeline()

figure_publication_time
```

Summary stats

```{r}
stats_pub_timeline(raw_hostmetagenome, raw_hostsinglegenome, raw_environmental) %>% 
  select(publication_doi) %>% 
  arrange %>% 
  distinct %>% 
  summarise(n = n())
```


# Cumulative Samples

```{r, fig.width=5, fig.height=7}

figure_cumulative_samples <- stats_cumulative_timeline(raw_hostmetagenome, raw_hostsinglegenome, raw_environmental) %>% plot_cumulative_timeline()

figure_cumulative_samples
```

Summary stats

```{r}
stats_cumulative_timeline(raw_hostmetagenome, raw_hostsinglegenome, raw_environmental) %>%
  group_by(List) %>%
  summarise(total = sum(count))
```


# Timelines Combined

```{r fig.height=7, fig.width=8}
figure_timelines <- figure_publication_time + figure_cumulative_samples + plot_layout(ncol = 2) + plot_annotation(tag_levels = 'a')

figure_timelines

ggsave("AncientMetagenomeDir-PublicationSample_Timeline.pdf",
       figure_timelines,
       device = cairo_pdf(),
       units = "in",
       width = 3.5,
       height = 4,
       scale = 2
       )

ggsave("AncientMetagenomeDir-PublicationSample_Timeline.png",
       figure_timelines,
       device = "png",
       units = "in",
       width = 3.5,
       height = 4,
       scale = 2
       )
```



# Geographic Spread

With known coordinates!

```{r, fig.height=7.5, fig.width=5}
data_map <- stats_map(raw_hostmetagenome, raw_hostsinglegenome, raw_environmental)

figure_map <- plot_map(data_map)

figure_map

```

How many countries does this cover?

```{r}
stats_map(raw_hostmetagenome, raw_hostsinglegenome, raw_environmental) %>% 
  ungroup() %>%
  select(geo_loc_name) %>% 
  distinct() %>% 
  summarise(n = n())
```

# Age Timeline

```{r}
# Combine data
dat_hist <- stats_age_timeline(raw_hostmetagenome,
                               raw_hostsinglegenome,
                               raw_environmental) #%>%
  #filter(sample_age < 50000)

# Plot the histograms - generating eachone separately as each need their own 
## custom binning mechanism to ensure readable plots
nclass <- "FD"
p1 <- plot_figure_age_timeline(dat_hist, dataset = "Host Associated Metagenome", nclass = nclass, log = TRUE, group_order = names(dir_colours))
p2 <- plot_figure_age_timeline(dat_hist, dataset = "Host Associated Single Genome", nclass = nclass, log = TRUE, group_order = names(dir_colours))
p3 <- plot_figure_age_timeline(dat_hist, dataset = "Environmental Metagenome", nclass = nclass, log = TRUE, group_order = names(dir_colours))

# Combine the histograms
figure_age_timeline <- ggplot() +
  geom_bar(stat = "identity", data = p1$data, aes(breaks, counts, fill = List),width = unique(p1$data$width), position = position_nudge(x = unique(p1$data$nudge))) +
  geom_bar(stat = "identity", data = p2$data, aes(breaks, counts, fill = List), width = unique(p2$data$width), position = position_nudge(x = unique(p2$data$nudge))) +
  geom_bar(stat = "identity", data = p3$data, aes(breaks, counts, fill = List), width = unique(p3$data$width), position = position_nudge(x = unique(p3$data$nudge))) +
  # scale_x_reverse() +
  scale_y_sqrt() +
  scale_x_reverse(breaks = c(2,3,4,5), labels = scales::comma(c(10^2, 10^3, 10^4, 10^5)), limits = c(5, 2)) +
  theme_classic() +
  scale_fill_manual(values = dir_colours, guide = guide_legend(ncol = 1, title = ""), breaks = names(dir_colours)) +
  ylab("Samples (n)") +
  xlab("Years before present")  +
  labs(fill = NULL) +
  theme(legend.position = "bottom") +
  facet_wrap(~factor(List, levels = names(dir_colours)), ncol = 1)

figure_age_timeline
```

# Time and Space Combined

```{r, fig.height=7, fig.width=7}
figure_map_age <- figure_map + figure_age_timeline + plot_annotation(tag_levels = "a")

figure_map_age

ggsave("AncientMetagenomeDir-Sample_Map_AgeTimeline.pdf",
       figure_map_age,
       device = cairo_pdf(),
       units = "in",
       width = 7,
       height = 7
       )

ggsave("AncientMetagenomeDir-Sample_Map_AgeTimeline.png",
       figure_map_age,
       device = "png",
       units = "in",
       width = 7,
       height = 7
       )
```


