---
title: "AncientMetagenomeDir - Analysis"
output: html_notebook
---

# Introduction

AncientMetagenomeDir is a repository listing published ancient metagenome (and 
related) samples. It contains critical metadata for ancient metagenome studies, 
and acts as a 'sign post' towards important data to help facilitate more robust
and efficient comparative data.

This page provides summary statistics of the current status of the directory.

# Libraries and Versions

```{r}
library(tidyverse)
library(scales)
library(lubridate)
library(maps)
library(patchwork)
sessionInfo()
```

# Preparation

## Functions

We will prepare some custom functions


### Loading

```{r}
## Load and standardise date across tables
load_thedir_data <- function(path, name) {
  read_tsv(path, col_types = cols()) %>%
    mutate(List = name) %>%
    select(List, everything())
}

```

### Publication Timeline

```{r}

stats_pub_timeline <- function(...) {
  x <- list(...)

  ## Get only relevent columns
  lapply(x, FUN = function(y) {
    select(
      y,
      List, publication_doi,
      publication_year
    ) %>% distinct()
  }) %>%
    bind_rows() %>%
    mutate(List = factor(List, levels = names(colours)))
}

plot_pub_timeline <- function(dat) {
  ## Get range so we plot x-axis nicely
  spanning_years <- dat %>% summarise(min = min(publication_year), max = max(publication_year))

  ggplot(dat, aes(publication_year, fill = List)) +
    scale_fill_manual(values = colours, guide = guide_legend(ncol = 1)) +
    scale_y_continuous(labels = scales::number_format(accuracy = 1)) +
    ylab("Number of publications") +
    xlab("Publication year") +
    geom_bar(bins = spanning_years$max - spanning_years$min, binwidth = 1) +
    scale_x_continuous(breaks = seq(spanning_years$min, spanning_years$max, 2)) +
    theme_classic() +
    theme(legend.position = "bottom") +
    facet_wrap(~List, ncol = 1) +
    labs(fill = NULL)
}
```

### Cumulative Samples Timeline

```{r}
stats_cumulative_timeline <- function(...) {
  ## Takes a list of AncientMetagenomeDir TSVs
  x <- list(...)

  ## Get only relevent columns
  dat <- lapply(x, FUN = function(y) {
    select(
      y,
      List, sample_name, publication_year,
      publication_year
    ) %>% distinct()
  }) %>%
    bind_rows() %>%
    mutate(List = factor(List, levels = names(colours)))

  spanning_years <- dat %>%
    ungroup() %>%
    summarise(min = min(publication_year), max = max(publication_year))

  ## Make fake base table to ensure all years for all lists
  ## Currently manually defined
  base_table <- list(
    seq(spanning_years$min, spanning_years$max, 1),
    seq(spanning_years$min, spanning_years$max, 1),
    seq(spanning_years$min, spanning_years$max, 1)
  )

  names(base_table) <- levels(dat$List)

  base_table <- base_table %>%
    enframe(name = "List", value = "publication_year") %>%
    unnest(publication_year)

  dat <- dat %>%
    group_by(List, publication_year) %>%
    summarise(count = n())

  dat %>%
    right_join(base_table, by = c("List", "publication_year")) %>%
    replace_na(list(count = 0)) %>%
    arrange(List, publication_year) %>%
    mutate(List = factor(List, levels = names(colours))) %>%
    group_by(List) %>%
    mutate(cumulative_sum = cumsum(count))
}


plot_cumulative_timeline <- function(x) {
  
  spanning_years <- list(min_year = min(x$publication_year), max_year = max(x$publication_year))

  ## Get range so we plot x-axis nicely
  ggplot(x, aes(publication_year, cumulative_sum, fill = List)) +
    geom_bar(stat = "identity") +
    scale_y_continuous(labels = scales::number_format(accuracy = 1)) +
    scale_x_continuous(breaks = seq(spanning_years$min_year, spanning_years$max_year, 2)) +
    theme_classic() +
    xlab("Publication year") +
    ylab("Number of samples (cumulative sum)") +
    scale_fill_manual(values = colours, guide = guide_legend(ncol = 1)) +
    theme(legend.position = "bottom") +
    facet_wrap(~List, ncol = 1) +
    labs(fill = NULL)
  
}


```


## Data Loading

We will load and analyse each list separately, as they contain slightly different
metadata depending on context.

```{r}
raw_hostmetagenome <- load_thedir_data("../../ancientmetagenome-hostassociated/ancientmetagenome-hostassociated.tsv", "Host Associated Metagenome") 
raw_hostsinglegenome <- load_thedir_data("../../ancientsinglegenome-hostassociated/ancientsinglegenome-hostassociated.tsv", "Host Associated Single Genome") 
raw_environmental <- load_thedir_data("../../ancientmetagenome-environmental/ancientmetagenome-environmental.tsv", "Environmental Metagenome") 
#raw_anthropogenic <- load_thedir_data("../../ancientmetagenome-anthropogenic/ancientmetagenome-anthropogenic.tsv", "Anthropogenic Metagenome") 
```

## Design Assets

```{r}
colours <- c(`Host Associated Metagenome` = "#73cff3",
            `Host Associated Single Genome` = "#d74182",
            `Environmental Metagenome` = "#2da46a")

notused_colours <- c(`Anthropogenic Metagenome` = "#d74182")

```


# Publication Timelines

```{r, fig.width=5, fig.height=7}
figure_publication_time <- stats_pub_timeline(raw_hostmetagenome, raw_hostsinglegenome, raw_environmental) %>% plot_pub_timeline()

figure_publication_time
```

Summary stats

```{r}
stats_pub_timeline(raw_hostmetagenome, raw_hostsinglegenome, raw_environmental) %>% 
  select(publication_doi) %>% 
  arrange %>% 
  distinct %>% 
  summarise(n = n())
```


# Cumulative Samples

```{r, fig.width=5, fig.height=7}

figure_cumulative_samples <- stats_cumulative_timeline(raw_hostmetagenome, raw_hostsinglegenome, raw_environmental) %>% plot_cumulative_timeline()

figure_cumulative_samples
```

Summary stats

```{r}
stats_cumulative_timeline(raw_hostmetagenome, raw_hostsinglegenome, raw_environmental) %>%
  group_by(List) %>%
  summarise(total = sum(count))
```


# Timelines Combined

```{r fig.height=7, fig.width=8}
figure_timelines <- figure_publication_time + figure_cumulative_samples + plot_layout(ncol = 2) + plot_annotation(tag_levels = 'a')

figure_timelines

ggsave("AncientMetagenomeDir-PublicationSample_Timeline.pdf",
       figure_timelines,
       device = cairo_pdf(),
       units = "in",
       width = 3.5,
       height = 4,
       scale = 2
       )

ggsave("AncientMetagenomeDir-PublicationSample_Timeline.png",
       figure_timelines,
       device = "png",
       units = "in",
       width = 3.5,
       height = 4,
       scale = 2
       )
```



# Geographic Spread

With known coordinates!

```{r eval = F}
world_map <- map_data("world")

map_hostmetagenome <- raw_hostmetagenome %>% 
  filter(!is.na(longitude)) %>% 
  group_by(latitude, longitude, community_type) %>% 
  summarise(Count = n()) %>%
  mutate(group = maps::map.where(database = "world", longitude, latitude))

## Find country name in maps::world a site is in based on long/lat

## Per Lat/Lon  
ggplot() +
  geom_polygon(data = world_map, aes(x = long, y = lat, group = group), fill = "white", colour = "grey") +
  geom_point(data = map_hostmetagenome, aes(x = longitude, y = latitude, fill = community_type, size = Count), shape = 21, alpha = 0.5) +
  scale_fill_brewer(palette ="Set1") +
  theme_linedraw()

## Country Heatmap
map_heatmapdata <- map_hostmetagenome %>% 
  group_by(community_type, group) %>% 
  summarise(Total = sum(Count))

world2 <-left_join(world, map_heatmapdata, by = c("name" = "group"))

ggplot() +
  geom_polygon(data = world_map, aes(x = long, y = lat, group = group), fill = "white", colour = "grey") +
  geom_polygon(data = map_heatmapdata, aes(fill = Total)) +
  scale_fill_brewer(palette ="Set1") +
  theme_minimal()

```

