---
title: "AncientMetagenomeDir - Analysis"
output: html_notebook
---

# Introduction

AncientMetagenomeDir is a repository listing published ancient metagenome (and 
related) samples. It contains critical metadata for ancient metagenome studies, 
and acts as a 'sign post' towards important data to help facilitate more robust
and efficient comparative data.

This page provides summary statistics of the current status of the directory.

# Libraries and Versions

```{r}
library(tidyverse)
library(scales)
library(lubridate)
library(maps)
library(patchwork)
sessionInfo()
```

# Preparation

## Functions

We will prepare some custom functions


### Loading

```{r}
## Load and standardise date across tables
load_thedir_data <- function(path, name) {
  read_tsv(path, col_types = cols()) %>%
    mutate(List = name) %>%
    select(List, everything())
}

```

### Publication Timeline

```{r}

stats_pub_timeline <- function(...) {
  x <- list(...)

  ## Get only relevent columns
  lapply(x, FUN = function(y) {
    select(
      y,
      List, publication_doi,
      publication_year
    ) %>% distinct()
  }) %>%
    bind_rows() %>%
    mutate(List = factor(List, levels = names(dir_colours)))
}

plot_pub_timeline <- function(dat) {
  ## Get range so we plot x-axis nicely
  spanning_years <- dat %>% summarise(min = min(publication_year), max = max(publication_year))

  ggplot(dat, aes(publication_year, fill = List)) +
    scale_fill_manual(values = dir_colours, guide = guide_legend(ncol = 1)) +
    scale_y_continuous(labels = scales::number_format(accuracy = 1)) +
    ylab("Number of publications") +
    xlab("Publication year") +
    geom_bar(bins = spanning_years$max - spanning_years$min, binwidth = 1) +
    scale_x_continuous(breaks = seq(spanning_years$min, spanning_years$max, 2)) +
    theme_classic() +
    theme(legend.position = "none",
          strip.background = element_blank()) +
    facet_wrap(~List, ncol = 1) +
    labs(fill = NULL)
}
```

### Cumulative Samples Timeline

```{r}
stats_cumulative_timeline <- function(...) {
  ## Takes a list of AncientMetagenomeDir TSVs
  x <- list(...)

  ## Get only relevent columns
  dat <- lapply(x, FUN = function(y) {
    select(
      y,
      List, sample_name, publication_year,
      publication_year
    ) %>% distinct()
  }) %>%
    bind_rows() %>%
    mutate(List = factor(List, levels = names(dir_colours)))

  spanning_years <- dat %>%
    ungroup() %>%
    summarise(min = min(publication_year), max = max(publication_year))

  ## Make fake base table to ensure all years for all lists
  ## Currently manually defined
  base_table <- list(
    seq(spanning_years$min, spanning_years$max, 1),
    seq(spanning_years$min, spanning_years$max, 1),
    seq(spanning_years$min, spanning_years$max, 1)
  )

  names(base_table) <- levels(dat$List)

  base_table <- base_table %>%
    enframe(name = "List", value = "publication_year") %>%
    unnest(publication_year)

  dat <- dat %>%
    group_by(List, publication_year) %>%
    summarise(count = n())

  dat %>%
    right_join(base_table, by = c("List", "publication_year")) %>%
    replace_na(list(count = 0)) %>%
    arrange(List, publication_year) %>%
    mutate(List = factor(List, levels = names(dir_colours))) %>%
    group_by(List) %>%
    mutate(cumulative_sum = cumsum(count))
}


plot_cumulative_timeline <- function(x) {
  
  spanning_years <- list(min_year = min(x$publication_year), max_year = max(x$publication_year))

  ## Get range so we plot x-axis nicely
  ggplot(x, aes(publication_year, cumulative_sum, fill = List)) +
    geom_bar(stat = "identity") +
    scale_y_continuous(labels = scales::number_format(accuracy = 1)) +
    scale_x_continuous(breaks = seq(spanning_years$min_year, spanning_years$max_year, 2)) +
    theme_classic() +
    xlab("Publication year") +
    ylab("Number of samples (cumulative sum)") +
    scale_fill_manual(values = dir_colours, guide = guide_legend(ncol = 1)) +
    theme(legend.position = "none",
          strip.background = element_blank()) +
    facet_wrap(~List, ncol = 1) +
    labs(fill = NULL)
  
}


```

### Map

```{r}
stats_map <- function(...){

x <- list(...)

dat <- lapply(x, FUN = function(y) {
  select(
    y,
    List, sample_name, geo_loc_name, latitude, longitude,
    publication_year
  ) %>% distinct()
}) %>%
  bind_rows() %>%
  mutate(List = factor(List, levels = names(dir_colours)))

dat <- dat %>%
  group_by(List, geo_loc_name, latitude, longitude) %>%
  summarise(count = n()) %>%
  mutate(List = factor(List, levels = names(dir_colours)))
}


plot_map <- function(dat){
  
  world_map <- map_data("world")
  
  ggplot() +
    geom_polygon(data = world_map, aes(x = long, y = lat, group = group), fill = "white", colour = "grey") +
    geom_point(data = dat, aes(x = longitude, y = latitude, fill = List, size = count), shape = 21, alpha = 0.5) +
    theme_linedraw() +
    facet_wrap(~List, ncol = 1) +
    theme_classic(base_size = 8) +
    scale_fill_manual(values = dir_colours, guide = FALSE) +
    theme(legend.position = "bottom", legend.direction="vertical",
          strip.background = element_blank()) +
    labs(fill = "Sample Type", size = "Sample Count") +
    xlab("Longitude") +
    ylab("Latitude") +
    guides(size = guide_legend(nrow = 1)) +
    xlim(-180, 180)
  
}
```


### Sample Age Timeline

```{r}
## from https://stackoverflow.com/a/11054781
reverselog_trans <- function(base = exp(1)) {
  trans <- function(x) -log(x, base)
  inv <- function(x) base^(-x)
  trans_new(paste0("reverselog-", format(base)), trans, inv,
    log_breaks(base = base),
    domain = c(1e-100, Inf)
  )
}


stats_age_timeline <- function(...){
  x <- list(...)
  
  dat <- lapply(x, FUN = function(y) {
    select(
      y,
      List, sample_name, geo_loc_name, latitude, longitude, sample_age, publication_year
    ) %>%
      distinct()
  }) %>%
    bind_rows() %>%
    mutate(List = factor(List, levels = names(dir_colours)))
  
  
  dat_hist <- dat %>%
    group_by(List, sample_age) %>%
    summarise(count = n())
}

plot_age_timeline <- function(dat){
  ggplot(dat, aes(sample_age, fill = List)) +
    geom_histogram(bins = 50) +
    scale_y_log10() +
    scale_x_reverse(labels = comma) +
    theme_classic(base_size = 8) +
    scale_fill_manual(values = dir_colours, guide = guide_legend(ncol = 1)) +
    ylab("Samples (n)") +
    xlab("Years before present")  +
    theme(legend.position = "none",
          strip.background = element_blank()) +
    facet_wrap(~List, nrow = 1)
}

```

## Data Loading

We will load and analyse each list separately, as they contain slightly different
metadata depending on context.

```{r}
raw_hostmetagenome <- load_thedir_data("../../ancientmetagenome-hostassociated/ancientmetagenome-hostassociated.tsv", "Host Associated Metagenome") 
raw_hostsinglegenome <- load_thedir_data("../../ancientsinglegenome-hostassociated/ancientsinglegenome-hostassociated.tsv", "Host Associated Single Genome") 
raw_environmental <- load_thedir_data("../../ancientmetagenome-environmental/ancientmetagenome-environmental.tsv", "Environmental Metagenome") 
#raw_anthropogenic <- load_thedir_data("../../ancientmetagenome-anthropogenic/ancientmetagenome-anthropogenic.tsv", "Anthropogenic Metagenome") 
```

## Design Assets

```{r}
dir_colours <- c(`Host Associated Metagenome` = "#73cff3",
            `Host Associated Single Genome` = "#d74182",
            `Environmental Metagenome` = "#2da46a")

notused_colours <- c(`Anthropogenic Metagenome` = "#d74182")

```


# Publication Timelines

```{r, fig.width=5, fig.height=7}
figure_publication_time <- stats_pub_timeline(raw_hostmetagenome, raw_hostsinglegenome, raw_environmental) %>% plot_pub_timeline()

figure_publication_time
```

Summary stats

```{r}
stats_pub_timeline(raw_hostmetagenome, raw_hostsinglegenome, raw_environmental) %>% 
  select(publication_doi) %>% 
  arrange %>% 
  distinct %>% 
  summarise(n = n())
```


# Cumulative Samples

```{r, fig.width=5, fig.height=7}

figure_cumulative_samples <- stats_cumulative_timeline(raw_hostmetagenome, raw_hostsinglegenome, raw_environmental) %>% plot_cumulative_timeline()

figure_cumulative_samples
```

Summary stats

```{r}
stats_cumulative_timeline(raw_hostmetagenome, raw_hostsinglegenome, raw_environmental) %>%
  group_by(List) %>%
  summarise(total = sum(count))
```


# Timelines Combined

```{r fig.height=7, fig.width=8}
figure_timelines <- figure_publication_time + figure_cumulative_samples + plot_layout(ncol = 2) + plot_annotation(tag_levels = 'a')

figure_timelines

ggsave("AncientMetagenomeDir-PublicationSample_Timeline.pdf",
       figure_timelines,
       device = cairo_pdf(),
       units = "in",
       width = 3.5,
       height = 4,
       scale = 2
       )

ggsave("AncientMetagenomeDir-PublicationSample_Timeline.png",
       figure_timelines,
       device = "png",
       units = "in",
       width = 3.5,
       height = 4,
       scale = 2
       )
```



# Geographic Spread

With known coordinates!

```{r, fig.height=7.5, fig.width=5}
figure_map <- stats_map(raw_hostmetagenome, raw_hostsinglegenome, raw_environmental) %>% plot_map

figure_map

ggsave("AncientMetagenomeDir-Sample_Map.pdf",
       figure_map,
       device = cairo_pdf(),
       units = "in",
       width = 1,
       height = 2,
       scale = 4
       )

ggsave("AncientMetagenomeDir-Sample_Map.png",
       figure_map,
       device = "png",
       units = "in",
       width = 1,
       height = 2,
       scale = 4
       )

```

How many countries does this cover?

```{r}
stats_map(raw_hostmetagenome, raw_hostsinglegenome, raw_environmental) %>% 
  ungroup() %>%
  select(geo_loc_name) %>% 
  distinct() %>% 
  summarise(n = n())
```

# Sample Age Timeline

```{r, fig.height=2.5, fig.width=7}
figure_age_timeline <- stats_age_timeline(raw_hostmetagenome, raw_hostsinglegenome, raw_environmental) %>% filter(sample_age < 50000) %>% plot_age_timeline

figure_age_timeline

```

# Time and Space Combined

```{r, fig.height=9, fig.width=5}
figure_age_map <- figure_age_timeline + figure_map + plot_layout(ncol = 1, heights = c(2,8)) + plot_annotation(tag_levels = "a")

figure_age_map

# ggsave("AncientMetagenomeDir-Sample_Age_Map.pdf",
#        figure_age_map,
#        device = cairo_pdf(),
#        units = "in",
#        width = 1.5,
#        height = 2,
#        scale = 4
#        )
# 
# ggsave("AncientMetagenomeDir-Sample_Age_Map.png",
#        figure_age_map,
#        device = "png",
#        units = "in",
#        width = 1.5,
#        height = 2,
#        scale = 4
#        )
```

